---
title: "CS-4200 Final Project"
author: "Ryan McMurray"
format:
  revealjs:
    theme: moon
    slide-number: true
    width: 1600
    height: 900
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

## Project Goal

- Analyze U.S. milk production data  
- Build a full data pipeline  
- Apply SQL and Machine Learning  

## Data Pipeline

1. Collect data from NASS API  
2. Clean and format data  
3. Store in SQLite database  
4. Query using SQL  
5. Apply machine learning models  

## Data Collection

```{r}
#| label: setup
#| include: false
#| cache: true

library(rnassqs)
library(DBI)
library(RSQLite)
library(knitr)
library(kableExtra)
library(ggplot2)

nassqs_auth(key = "F69DFA8F-B2BA-3DBD-883E-C62A00818A29")
params <- list(
  commodity_desc = "MILK",
  statisticcat_desc = "PRODUCTION",
  unit_desc = "LB",
  agg_level_desc = "STATE",
  freq_desc = "ANNUAL",
  year__GE = 2010
)

raw_data <- nassqs(params)
```

## Data Cleaning
```{r}
#| label: data_cleaning
raw_data$Value <- gsub(",", "", raw_data$Value)
raw_data$Value <- suppressWarnings(as.numeric(raw_data$Value))

clean_data <- raw_data[!is.na(raw_data$Value), ]
clean_data <- subset(clean_data, unit_desc == "LB")

clean_data$year <- as.numeric(clean_data$year)
```

## SQL Database

```{r}
con <- dbConnect(SQLite(), ":memory:")
dbWriteTable(con, "milk_data", clean_data)
```

```{sql}
--| label: sql_filtering
--| connection: con
--| output.var: "filtered_milk"
--| echo: true

SELECT state_alpha, year, Value 
FROM milk_data 
ORDER BY Value DESC
LIMIT 5;
```

```{r}
#| label: display_results
#| echo: false

knitr::kable(filtered_milk) |>
  kable_styling(full_width = FALSE, font_size = 24)
```

## Top Producers

```{r}
#| label: bar_chart

top_summary <- aggregate(Value ~ state_alpha, data = clean_data, max)
top_summary <- top_summary[order(-top_summary$Value), ][1:5, ]

ggplot(top_summary, aes(x = reorder(state_alpha, Value), y = Value)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 5 Milk Producing States",
    x = "State",
    y = "Production (LB)"
  ) +
  theme_minimal()
```

## Linear Regression

```{r}
#| label: linear_regression
model <- lm(Value ~ year + state_alpha, data = clean_data)
summary(model)
```

## Production Trends

```{r}
#| label: linear_plot
library(ggplot2)

clean_data$year <- as.numeric(clean_data$year)

top_states <- names(sort(
  tapply(clean_data$Value, clean_data$state_alpha, max),
  decreasing = TRUE
))[1:5]

plot_data <- subset(clean_data, state_alpha %in% top_states)

ggplot(plot_data, aes(x = year, y = Value, color = state_alpha)) +
  geom_line(linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Milk Production Trends (Top 5 States)",
    x = "Year",
    y = "Production (LB)",
    color = "State"
  ) +
  theme_minimal()
```

## K-Means

```{r}
#| label: k-means

kmeans_data <- clean_data[, c("year", "Value")]
scaled_data <- scale(kmeans_data)
scaled_data <- scaled_data[complete.cases(scaled_data), ]

set.seed(123)
kmeans_result <- kmeans(scaled_data, centers = 3)

valid_rows <- complete.cases(kmeans_data)
clean_data$cluster <- NA
clean_data$cluster[valid_rows] <- kmeans_result$cluster
```

## Clustering Results

```{r}
ggplot(
  clean_data[!is.na(clean_data$cluster), ],
  aes(x = year, y = Value, color = factor(cluster))
) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Milk Production Clusters",
    x = "Year",
    y = "Production (LB)",
    color = "Cluster"
  ) +
  theme_minimal()
```

## Key Insights

- Production is concentrated in a few states  
- Trends are stable over time  
- Clustering reveals natural groupings  
- Data cleaning is critical for analysis  

## Conclusion

- Built a full data pipeline  
- Combined SQL and Machine Learning  
- Demonstrated Big Data workflow  