---
title: "CS-4200 Final Project"
author: "Ryan McMurray"
format:
  pdf:
    documentclass: article
    geometry:
      - margin=1in
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Introduction

This project demonstrates a complete data science workflow using the Titanic dataset, with a focus on exploratory analysis, predictive modeling, and classification.

The analysis follows a structured pipeline:

- Data preprocessing
  - Cleaning missing values and formatting variables for modeling
- Exploratory Data Analysis (EDA)
  - Visualizing relationships between survival, class, age, and fare
- Predictive modeling
  - Linear and nonlinear regression to understand continuous relationships
- Classification
  - Comparing logistic regression, decision trees, and random forests to predict survival outcomes

## Clean

```{r}
#| label: load_and_clean
#| echo: true
library(titanic)
library(tidyverse)

df_clean <- titanic_train |>
  tibble() |>
  select(Survived, Pclass, Sex, Age, Fare) |>
  drop_na() |>
  distinct() |>
  mutate(
    Survived = factor(Survived, levels = c(0, 1)),
    Sex = as.factor(Sex),
    Pclass = as.factor(Pclass),
    Age = as.numeric(Age),
    Fare = as.numeric(Fare)
  )

summary(df_clean$Age)
summary(df_clean$Fare)
```

## EDA

```{r}
#| label: viz_dashboard
#| echo: false
library(patchwork)
library(ggplot2)
library(scales)

p1 <- ggplot(df_clean, aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Survival Rate by Class") +
  theme_minimal()

p2 <- ggplot(df_clean, aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "dodge") +
  labs(title = "Survival Counts by Class") +
  theme_minimal()

p3 <- ggplot(df_clean, aes(x = Survived, y = Fare, fill = Survived)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(title = "Fare Distribution") +
  theme_minimal()

p4 <- ggplot(df_clean, aes(x = Age, fill = Survived)) +
  geom_density(alpha = 0.4) +
  labs(title = "Age Distribution") +
  theme_minimal()

p5 <- df_clean |>
  filter(Pclass == 1) |>
  ggplot(aes(x = Fare, fill = Survived)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  scale_x_log10() +
  labs(title = "First Class Fare") +
  theme_minimal()

(p1 | p2) / (p3 | p4) / p5
```

## Regression Comparison Predicting Fare Price

```{r}
#| label: model_comparison_all
library(tidyverse)
library(rpart)
library(randomForest)

set.seed(4200)

# ---- CLEANING FOR MODELING ----
df_model <- df_clean |>
  mutate(
    Fare = log1p(Fare), # FIX: stabilizes skew
    Sex = as.factor(Sex),
    Pclass = as.factor(Pclass)
  ) |>
  drop_na()

# ---- TRAIN/TEST SPLIT ----
train_index <- sample(seq_len(nrow(df_model)), 0.8 * nrow(df_model))
train <- df_model[train_index, ]
test <- df_model[-train_index, ]

# ---- MODELS ----

# Linear Regression (baseline)
lin_model <- lm(Fare ~ Age + Pclass + Sex, data = train)

# Decision Tree (pruned-ish behavior via minsplit)
tree_model <- rpart(
  Fare ~ Age + Pclass + Sex,
  data = train,
  method = "anova",
  control = rpart.control(minsplit = 20, cp = 0.01)
)

# Random Forest (stronger defaults)
rf_model <- randomForest(
  Fare ~ Age + Pclass + Sex,
  data = train,
  ntree = 500,
  mtry = 2,
  importance = TRUE
)

# Polynomial regression (captures curvature)
poly_model <- lm(
  Fare ~ poly(Age, 2) + Pclass + Sex,
  data = train
)

# ---- METRICS ----
rmse <- function(a, p) sqrt(mean((a - p)^2))
r2 <- function(a, p) 1 - sum((a - p)^2) / sum((a - mean(a))^2)

# ---- PREDICTIONS ----
pred_lin <- predict(lin_model, test)
pred_tree <- predict(tree_model, test)
pred_rf <- predict(rf_model, test)
pred_poly <- predict(poly_model, test)

# ---- RESULTS TABLE ----
results <- tibble(
  Model = c("Linear", "Decision Tree", "Random Forest", "Polynomial (deg 2)"),
  RMSE = c(
    rmse(test$Fare, pred_lin),
    rmse(test$Fare, pred_tree),
    rmse(test$Fare, pred_rf),
    rmse(test$Fare, pred_poly)
  ),
  R2 = c(
    r2(test$Fare, pred_lin),
    r2(test$Fare, pred_tree),
    r2(test$Fare, pred_rf),
    r2(test$Fare, pred_poly)
  )
)

results
```

Random Forest performs best overall, with the lowest error (RMSE = 0.634) and highest explanatory power (R² = 0.604).
Decision Tree is second-best, slightly improving over linear regression (R² = 0.577).
Polynomial regression provides only minor improvement over linear regression, suggesting limited nonlinear structure beyond what simpler models capture.
Linear regression performs worst, confirming that the relationship between variables is not fully linear.

## Side-by-side Regression Comparison

```{r}
#| label: model_comparison_plot

library(tidyverse)
library(patchwork)

# Ensure consistent ordering (important for readability)
results <- results |>
  mutate(Model = factor(Model, levels = Model))

# RMSE plot (lower is better)
p_rmse <- ggplot(results, aes(x = Model, y = RMSE, fill = Model)) +
  geom_col(show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "Model Error (RMSE)",
    x = NULL,
    y = "RMSE"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# R² plot
p_r2 <- ggplot(results, aes(x = Model, y = R2, fill = Model)) +
  geom_col(show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "Model Fit (R²)",
    x = NULL,
    y = "R²"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Combine side-by-side
p_rmse + p_r2
```

# Categorization

## Classification Models

```{r}
#| label: models
library(randomForest)
library(rpart)

set.seed(4200)

train_index <- sample(seq_len(nrow(df_clean)), 0.8 * nrow(df_clean))
train <- df_clean[train_index, ]
test <- df_clean[-train_index, ]

# Logistic Regression
log_model <- glm(
  Survived ~ Age + Fare + Sex + Pclass,
  data = train,
  family = binomial
)

# Decision Tree
tree_model <- rpart(
  Survived ~ Age + Fare + Sex + Pclass,
  data = train,
  method = "class"
)

# Random Forest
rf_model <- randomForest(
  Survived ~ Age + Fare + Sex + Pclass,
  data = train
)

pred_log <- ifelse(predict(log_model, test, type = "response") > 0.5, 1, 0)
pred_tree <- predict(tree_model, test, type = "class")
pred_rf <- predict(rf_model, test)

accuracy <- function(pred, actual) mean(pred == actual)

results_cls <- tibble(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    accuracy(pred_log, test$Survived),
    accuracy(pred_tree, test$Survived),
    accuracy(pred_rf, test$Survived)
  )
)

ggplot(results_cls, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col(show.legend = FALSE) +
  theme_minimal() +
  labs(
    x = NULL,
    y = "Accuracy"
  )
```

## Side-by-side Classification Comparison for Survival

```{r}
#| label: classification_metrics
library(dplyr)
library(tibble)

# Helper: confusion matrix metrics for binary classification
metrics <- function(actual, pred) {
  cm <- table(Predicted = pred, Actual = actual)

  TP <- cm["1", "1"]
  TN <- cm["0", "0"]
  FP <- cm["1", "0"]
  FN <- cm["0", "1"]

  accuracy <- (TP + TN) / sum(cm)
  precision <- TP / (TP + FP)
  recall <- TP / (TP + FN)
  f1 <- 2 * precision * recall / (precision + recall)
  specificity <- TN / (TN + FP)

  tibble(
    Accuracy = accuracy,
    Precision = precision,
    Recall = recall,
    F1 = f1,
    Specificity = specificity
  )
}

# Predictions (assumes you already defined these)
log_metrics <- metrics(test$Survived, pred_log)
tree_metrics <- metrics(test$Survived, pred_tree)
rf_metrics <- metrics(test$Survived, pred_rf)

results_class <- bind_rows(
  cbind(Model = "Logistic Regression", log_metrics),
  cbind(Model = "Decision Tree", tree_metrics),
  cbind(Model = "Random Forest", rf_metrics)
)

results_class
```

- Random Forest performs best overall
  - Highest Accuracy (0.839) and F1 score (0.803)
  - Strong balance of Precision (0.839) and Recall (0.770)
  - Good generalization across both classes
- Decision Tree is second best
  - Highest Precision (0.872), meaning fewer false positives
  - Lower Recall (0.672), so it misses more survivors
- Logistic Regression is the baseline model
  - More balanced but weaker overall performance
  - Lower ability to capture nonlinear relationships

## Random Forest Confusion Matrix

```{r}
cm <- table(Predicted = pred_rf, Actual = test$Survived)
cm_df <- as.data.frame(cm)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 5) +
  scale_fill_gradient(low = "steelblue", high = "darkred") +
  theme_minimal() +
  labs(
    x = "Actual",
    y = "Predicted"
  )
```

## Key Findings

- Passenger class strongly impacts survival
- Clustering reveals distinct passenger groups
- Random Forest performs best among models
- Fare distribution is highly skewed (log scaling required)

## Conclusion

This project demonstrates:

- Data cleaning (Chapter 9)
- Clustering analysis (Chapter 10)
- Classification comparison (Chapter 11)

The Titanic dataset clearly shows how socioeconomic factors influenced survival.